Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-05-01 papers/halfdisk/halfdisk.tex
Further edits

\newcommand{\half}{\frac{1}{2}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\No}{\mathbb{N}_0}
\newcommand{\Ylm}{Y^m_l}
\newcommand{\Ylmfull}{Y^m_l(\theta,\varphi)}
\newcommand{\Plm}{P^m_l}
\newcommand{\costheta}{\cos\theta}
\newcommand{\sintheta}{\sin\theta}
\newcommand{\cosphi}{\cos\varphi}
\newcommand{\sinphi}{\sin\varphi}
\newcommand{\eimphi}{e^{im\varphi}}
\newcommand{\alphalm}{\alpha^m_l}
\newcommand{\clm}{c^m_l}
\newcommand{\ctilde}{\tilde{c}^m_l}
\newcommand{\ctildemod}{\tilde{c}^{|m|}_l}
\newcommand{\chat}{\hat{c}^m_l}
\newcommand{\chatmod}{\hat{c}^{|m|}_l}
\newcommand{\ddx}{\frac{\mathrm{d}}{\mathrm{d}x}}
\newcommand{\pddx}{\frac{\partial}{\partial x}}
\newcommand{\pddy}{\frac{\partial}{\partial y}}
\newcommand{\pddn}{\frac{\partial}{\partial n}}
\newcommand{\dmdxm}{\frac{\mathrm{d}^m}{\mathrm{d}x^m}}

@@

\begin{proof}
We proceed with the case for the operator $D_x^{(a,b)}$ for partial differentiation by $x$. Since $\{\hdop^{(a+1, b+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddx \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: \hdop^{(a+1, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We can use an integration-by-parts argument to show that the only non-zero coefficients of this expansion are when $m = n-1, n-2$, $j = k, k-2$ and $0 \le j \le m$. First, note that
\begin{align*}
c_{m,j} = \frac{\ip<\pddx \hdopnkab, \hdop^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}}}{\norm{\hdop^{(a+1, b+1)}_{m,j}}^2_{W^{(a+1, b+1)}}}.
\end{align*}
Second, note that
\begin{align}
	\rho'(x) = -x \: \rho(x)^{-2}. \label{eqn:rhoderivative}
\end{align}
Now, using a change of variables and (\ref{eqn:rhoderivative}), we have that
\begin{align*}
	&\ip<\pddx \hdopnkab, \hdop^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}} \nonumber \\ 
	&= \ip< x \: \genjacnmk^{(a, b+k+\half) \: \prime} (x), \: \rho(x)^{k+j+2} \: \genjacmmj^{(a, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} 
		\: \ip< P_k^{(b,b)}, \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  \nonumber \\ 
	& \quad - k \: \ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} 
		\: \ip< P_k^{(b,b)}, \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  \nonumber \\ 
	& \quad + \: \ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} 
		\: \ip< y \: P_k^{(b,b) \: \prime} (y), \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  
\end{align*}
We can see that, from each term's second factor as written, the above is zero for $j < k-2$. Now, using integration-by-parts and noting that the weight $\genjacw^{(a,b)}$ is a polynomial degree $a+2b$ and vanishes at the limits of the integral for positive parameters $a,b$, we have that
\begin{align*}
	&\ip< x \: \genjacnmk^{(a, b+k+\half) \: \prime}, \: \rho(x)^{k+j+2} \: \genjacmmj^{(a, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}}  \\
	&= \ip<\genjacnmk^{(a, b+k+\half) \: \prime}, \: \genjacw^{(a+1, b+1+\half(k+j+1))} \: \genjacmmj^{(a, b+1+j+\half)} >_1 \\
	&= \ip<\genjacnmk^{(a, b+k+\half)}, \: \ddx \Big[\genjacw^{(a+1, b+1+\half(k+j+1))} \: \genjacmmj^{(a, b+1+j+\half)} \Big] >_1 \\
	&= \ip<\genjacnmk^{(a, b+k+\half)}, \: \Big[(a+1) \: \genjacw^{(a, b+1+\half(k+j+1))} -2(b+1+\half(k+j+1)) \: \genjacw^{(a+2, b+\half(k+j+1))} \Big] \: \genjacmmj^{(a, b+1+j+\half)} >_1 \\
		&\quad \quad \quad + \ip<\genjacnmk^{(a, b+k+\half)}, \: \genjacw^{(a+1, b+1+\half(k+j+1))} \: \genjacmmj^{(a, b+1+j+\half) \: \prime}>_1 \\
	&= \ip<\genjacnmk^{(a, b+k+\half)}, \: \Big[(a+1) \: \genjacw^{(0,1+\half(j-k))} -2(b+1+\half(k+j+1)) \: \genjacw^{(2, \half(j-k))} \Big] \: \genjacmmj^{(a, b+1+j+\half)} >_{\genjacw^{(a, b+k+\half)}} \\
		&\quad \quad \quad + \ip<\genjacnmk^{(a, b+k+\half)}, \: \genjacw^{(1, 1+\half(j-k))} \: \genjacmmj^{(a, b+1+j+\half) \: \prime}>_{\genjacw^{(a, b+k+\half)}}
\end{align*}
which is zero for $m < n - 2$, and further that
\begin{align*}
	&\ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} \\
	&= \ip< \genjacnmk^{(a, b+k+\half)}(x), \: \genjacw^{(2, \half(j-k))} \: \genjacmmj^{(a, b+1+j+\half)} >_{\genjacw^{(a, b+k+\half)}}
\end{align*}
which is also zero for $m < n - 2$.

\bstodoinline{Finish this proof!!!}

For the operator for partial differentiation by $y$, we can also expand $\pddy \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: P^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. By a similar argument to before, we have that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. This is then calculated by taking inner products, i.e. 
\begin{align*}
c_{n-1,k-1} = \frac{\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{n-1,k-1} >_{W^{(a, b+1)}}}{\norm{\hdop^{(a, b+1)}_{n-1,k-1}}^2_{W^{(a, b+1)}}}
\end{align*}
where
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{n-1,k-1} >_{W^{(a, b+1)}} = \normgenjac^{(a, b+k+\half)} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}

We can gain the non-zero entries of the weighted differential operators similarly, by noting that
\begin{align}
	\pddx \Wab(x,y) &= a W^{(a-1, b)}(x,y) - 2bW^{(a+1, b-1)}(x,y) \label{eqn:weightderivativex} \\
	\pddy \Wab(x,y) &= -2b \: y \: W^{(a, b-1)}(x,y) \label{eqn:weightderivativey}
\end{align}
and also that
\begin{align*}
	\ip< \Wab \hdopnkab, W^{(\tilde{a},\tilde{b})} \hdopmj^{(\tilde{a},\tilde{b})} >_{W^{(-\tilde{a},-\tilde{b})}} = \ip< \hdopnkab, \hdopmj^{(\tilde{a},\tilde{b})} >_\Wab.
\end{align*}
\end{proof}

>>>>

\newcommand{\half}{\frac{1}{2}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\No}{\mathbb{N}_0}
\newcommand{\Ylm}{Y^m_l}
\newcommand{\Ylmfull}{Y^m_l(\theta,\varphi)}
\newcommand{\Plm}{P^m_l}
\newcommand{\costheta}{\cos\theta}
\newcommand{\sintheta}{\sin\theta}
\newcommand{\cosphi}{\cos\varphi}
\newcommand{\sinphi}{\sin\varphi}
\newcommand{\eimphi}{e^{im\varphi}}
\newcommand{\alphalm}{\alpha^m_l}
\newcommand{\clm}{c^m_l}
\newcommand{\ctilde}{\tilde{c}^m_l}
\newcommand{\ctildemod}{\tilde{c}^{|m|}_l}
\newcommand{\chat}{\hat{c}^m_l}
\newcommand{\chatmod}{\hat{c}^{|m|}_l}
\newcommand{\ddx}{\frac{\mathrm{d}}{\mathrm{d}x}}
\newcommand{\ddy}{\frac{\mathrm{d}}{\mathrm{d}y}}
\newcommand{\pddx}{\frac{\partial}{\partial x}}
\newcommand{\pddy}{\frac{\partial}{\partial y}}
\newcommand{\pddn}{\frac{\partial}{\partial n}}
\newcommand{\dmdxm}{\frac{\mathrm{d}^m}{\mathrm{d}x^m}}

@@

\begin{proof}
We proceed with the case for the operator $D_y^{(a,b)}$ for partial differentiation by $y$, we can also expand $\pddy \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: P^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We can use an integration-by-parts argument to show that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. First, note that
\begin{align*}
	c_{m,j} = \frac{\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}}}{\norm{\hdop^{(a, b+1)}_{m,j}}^2_{W^{(a, b+1)}}}.
\end{align*}
Then, using the change of variable $\tilde{y} = \frac{y}{\rho(x)}$, we have that
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}} &= \ip< \genjacnmk^{(a,b+k+\half)}, \rho(x)^{k+j} \: \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}} \\
	&= \ip< \genjacnmk^{(a,b+k+\half)}, \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1+\half(k+j))}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}
Now, integrating by parts, noting that the weight $w_P^{(b)}$ is a polynomial degree $2b$ and vanishes at the limits of the integral for positive parameter $b$, we have that
\begin{align*}
	\ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}} &= \ip< P_k^{(b,b) \: \prime}, \: w_P^{(b+1)} \: P_{j}^{(b+1,b+1)} >_1 \\
	&= \ip< P_k^{(b,b)}, \: \ddy [w_P^{(b+1)} \: P_{j}^{(b+1,b+1)}] >_1 \\
	&= \ip< P_k^{(b,b)}, \: P_{j}^{(b+1,b+1) \: \prime} >_{w_P^{(b+1)}} - 2 b \ip< P_k^{(b,b)}, \: y \: P_{j}^{(b+1,b+1)} >_{w_P^{(b)}} \\
	&= \ip< P_k^{(b,b)}, \: w_P^{(1)} \: P_{j}^{(b+1,b+1) \: \prime} -2b \:y \: P_{j}^{(b+1,b+1)} >_{w_P^{(b)}} 
\end{align*}
which is zero for $j < k-1$. Further,
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,k-1} >_{W^{(a, b+1)}} &= \ip< \genjacnmk^{(a,b+k+\half)}, \genjac_{m+1-k}^{(a,b+k+\half)} >_{\genjacw^{(a,b+k+\half)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}},
\end{align*}
which is zero for $m < n-1$. Finally, we simply have that
\begin{align*}
	c_{n-1,k-1} &= \frac{1}{\normjac^{(b)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}

We next consider the case for the operator $D_x^{(a,b)}$ for partial differentiation by $x$. Since $\{\hdop^{(a+1, b+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddx \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: \hdop^{(a+1, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. As before, we can use an integration-by-parts argument to show that the only non-zero coefficients of this expansion are when $m = n-1, n-2$, $j = k, k-1, k-2$ and $0 \le j \le m$. First, note that
\begin{align*}
c_{m,j} = \frac{\ip<\pddx \hdopnkab, \hdop^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}}}{\norm{\hdop^{(a+1, b+1)}_{m,j}}^2_{W^{(a+1, b+1)}}}.
\end{align*}
Second, note that
\begin{align}
	\rho'(x) = -x \: \rho(x)^{-2}. \label{eqn:rhoderivative}
\end{align}
Now, using the change of variable $\tilde{y} = \frac{y}{\rho(x)}$ and (\ref{eqn:rhoderivative}), we have that
\begin{align*}
	&\ip<\pddx \hdopnkab, \hdop^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}} \nonumber \\ 
	&= \ip< x \: \genjacnmk^{(a, b+k+\half) \: \prime} (x), \: \rho(x)^{k+j+2} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} 
		\: \ip< P_k^{(b,b)}, \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  \nonumber \\ 
	& \quad - k \: \ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} 
		\: \ip< P_k^{(b,b)}, \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  \nonumber \\ 
	& \quad + \: \ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} 
		\: \ip< y \: P_k^{(b,b) \: \prime} (y), \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  
\end{align*}
We can see that, from each term's second factor as written, the above is zero for $j < k-2$. Now, using integration-by-parts and noting that the weight $\genjacw^{(a,b)}$ is a polynomial degree $a+2b$ and vanishes at the limits of the integral for positive parameters $a,b$, we have that
\begin{align*}
	&\ip< x \: \genjacnmk^{(a, b+k+\half) \: \prime}, \: \rho(x)^{k+j+2} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}}  \\
	&= \ip<\genjacnmk^{(a, b+k+\half) \: \prime}, \: \genjacw^{(a+1, b+1+\half(k+j+1))} \: \genjacmmj^{(a+1, b+1+j+\half)} >_1 \\
	&= \ip<\genjacnmk^{(a, b+k+\half)}, \: \ddx \Big[\genjacw^{(a+1, b+1+\half(k+j+1))} \: \genjacmmj^{(a+1, b+1+j+\half)} \Big] >_1 \\
	&= \ip<\genjacnmk^{(a, b+k+\half)}, \: \Big[(a+1) \: \genjacw^{(a, b+1+\half(k+j+1))} -2(b+1+\half(k+j+1)) \: \genjacw^{(a+2, b+\half(k+j+1))} \Big] \: \genjacmmj^{(a, b+1+j+\half)} >_1 \\
		&\quad \quad \quad + \ip<\genjacnmk^{(a, b+k+\half)}, \: \genjacw^{(a+1, b+1+\half(k+j+1))} \: \genjacmmj^{(a+1, b+1+j+\half) \: \prime}>_1 \\
	&= \ip<\genjacnmk^{(a, b+k+\half)}, \: \Big[(a+1) \: \genjacw^{(0,1+\half(j-k))} -2(b+1+\half(k+j+1)) \: \genjacw^{(2, \half(j-k))} \Big] \: \genjacmmj^{(a, b+1+j+\half)} >_{\genjacw^{(a, b+k+\half)}} \\
		&\quad \quad \quad + \ip<\genjacnmk^{(a, b+k+\half)}, \: \genjacw^{(1, 1+\half(j-k))} \: \genjacmmj^{(a+1, b+1+j+\half) \: \prime}>_{\genjacw^{(a, b+k+\half)}}
\end{align*}
which is zero for $m < n - 2$, and further that
\begin{align*}
	&\ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} \\
	&= \ip< \genjacnmk^{(a, b+k+\half)}(x), \: \genjacw^{(2, \half(j-k))} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+k+\half)}}
\end{align*}
which is also zero for $m < n - 2$.

We can gain the non-zero entries of the weighted differential operators similarly, by noting that
\begin{align}
	\pddx \Wab(x,y) &= a W^{(a-1, b)}(x,y) - 2bW^{(a+1, b-1)}(x,y) \label{eqn:weightderivativex} \\
	\pddy \Wab(x,y) &= -2b \: y \: W^{(a, b-1)}(x,y) \label{eqn:weightderivativey}
\end{align}
and also that
\begin{align*}
	\ip< \Wab \hdopnkab, W^{(\tilde{a},\tilde{b})} \hdopmj^{(\tilde{a},\tilde{b})} >_{W^{(-\tilde{a},-\tilde{b})}} = \ip< \hdopnkab, \hdopmj^{(\tilde{a},\tilde{b})} >_\Wab.
\end{align*}
